A High Efficiency MAC Protocol for WLANs: Providing Fairness in Dense Scenarios

### 2015-10-22 relatedWork.tex
Second draft complete

The following are MAC protocols for WLANs, distributed and capable of attaining greater throughput than CSMA/CA by constructing a collision-free schedule.

@@

Zero Collision MAC (ZC-MAC)~\cite{ZMAC} achieves a zero collision schedule for WLANs in a fully distributed way. It does so by allowing contenders to reserve one empty slot from a predefined virtual schedule of $N$-slots in length. Backlogged stations pick a slot in the virtual cycle to attempt transmission. If two or more stations picked the same slot in the cycle, their transmissions will eventually collide. This forces the involved contenders to randomly and uniformly select other empty slot from those detected empty in the previous cycle plus the slot where they collided. When all $M$ stations reserve a different slot, a collision-free schedule is achieved.

ZC-MAC is able to outperform CSMA/CA under different scenarios. Nevertheless, given that the length of ZC MAC's virtual cycle has to be predefined without actual knowledge of the real number of contenders in the deployment, the protocol is unable to provide a collision-free schedule when $M>N$. Furthermore, if $N$ is overestimated ($N\gg M$), the fixed-width empty slots between each contender's successful transmission are no longer negligible and contribute to the degradation of the network performance\textcolor{red}{\footnote{\textcolor{red}{With a very large $N$ and $M\leq N$, the overall aggregated throughput will start to decrease as $M\rightarrow N$. This is due to the very big period of time between successful transmissions. For the opposite case, that is $M>N$, the same effect is produced albeit with an increased number of collisions.}}}. Additionally, Z-MAC nodes require common knowledge of where the virtual schedule starts/ends. This is not considered in CSMA/CA and constitutes an obstacle towards standardisation.

>>>>

\textcolor{red}{Performing time slot reservation for each transmission is a well known technique for increasing the throughput and mantanining Quality of Service (QoS) in TDMA schemes, like ATM~\cite{frigon2001dynamic}. Applying the same concept to CSMA networks by modifying DCF's random backoff proceedure provides similar benefits~\cite{HE}. The following are MAC protocols for WLANs, distributed and capable of attaining greater throughput than CSMA/CA by constructing collision-free schedules using reservation techniques.}

@@

Zero Collision MAC (ZC-MAC)~\cite{ZMAC} achieves a zero collision schedule for WLANs in a fully distributed way. It does so by allowing contenders to reserve one empty slot from a predefined virtual schedule of $M$-slots in length. Backlogged stations pick a slot in the virtual cycle to attempt transmission. If two or more stations picked the same slot in the cycle, their transmissions will eventually collide. This forces the involved contenders to randomly and uniformly select other empty slot from those detected empty in the previous cycle plus the slot where they collided. When all $N$ stations reserve a different slot, a collision-free schedule is achieved.

ZC-MAC is able to outperform CSMA/CA under different scenarios. Nevertheless, given that the length of ZC MAC's virtual cycle has to be predefined without actual knowledge of the real number of contenders in the deployment, the protocol is unable to provide a collision-free schedule when $N>M$. Furthermore, if $M$ is overestimated ($M\gg N$), the fixed-width empty slots between each contender's successful transmission are no longer negligible and contribute to the degradation of the network performance\textcolor{red}{\footnote{\textcolor{red}{With a very large $M$ and $N\ll M$, the overall aggregated throughput will start to decrease as $N\rightarrow M$. This is due to the very big period of time between the successful transmissions of each node. For the opposite case, that is $N>M$, the throughput degradation will be accelerated due to collisions.}}}. Additionally, Z-MAC nodes require common knowledge of where the virtual schedule starts/ends. This is not considered in CSMA/CA and constitutes an obstacle towards standardisation.

### 2015-10-22 results.tex
Second draft complete

\section{Results}\label{results}
	\begin{figure*}[tb]
		\centering
		\includegraphics[width=0.5\linewidth,angle=-90]{figures/tonFigs/saturation-combined.eps}
		\caption{Saturation traffic results: a) Throughput under saturated conditions; b) Average percentage of collision slots: fraction of time slots containing collisions; c) Evolution of the fraction of collision slots in a scenario with 70 saturated stations; d) Average time between successful transmissions (sx tx)}
		\label{fig:satResults}
	\end{figure*}
	%\begin{figure*}[tb]
%	\centering
%		\subcaptionbox{Throughput under saturated conditions\label{fig:throughput-sat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/throughput-saturated-w-BOS/throughput-saturated-w-BOS-TON.eps}}\hfill
%		\subcaptionbox{Average percentage of collision slots: the fraction of time slots containing collisions\label{fig:collisions-sat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/collisions-saturated/collisions-saturated-TON.eps}}\\
%		\subcaptionbox{Evolution of the fraction of collision slots in a scenario with 70 saturated stations\label{fig:collisions-evolution}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/slots/Pc-evolution-TON.eps}}\hfill
%		\subcaptionbox{Average time between successful transmissions: averaging over all stations\label{fig:serviceTime-sat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/timeBetweenSxTx-sat/timeBetweenSxTx-multiplot-sat-TON.eps}}\\
%		\caption{Saturated traffic results}
%		\label{fig:satTest}
%	\end{figure*}

@@

	This section aims at overviewing the throughput of CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ in saturation, as well as the collision probability, the average time between successful transmissions and the effect of clock drift over the throughput.
	\\
	\subsubsection{Throughput}
	CSMA/ECA$_{\text{Hys+FS}}$ nodes are able to build a collision-free schedule, use the channel more efficiently, and experience a throughput increase as seen in Figure~\ref{fig:satResults}a. Hysteresis allows the allocation of more contenders by increasing the length of the collision-free schedule, while Fair Share ensures an even distribution of the available throughput. This is reflected by the average backoff stage, which value increases with the number of contenders. In contrast, CSMA/CA throughput keeps decreasing due to an augmented number of collisions as the number of nodes increases (see Figure~\ref{fig:satResults}b). Further, Figure~\ref{fig:satResults}c shows the fraction of collision slots for CSMA/ECA$_{\text{Hys+FS}}$ and CSMA/CA as simulation time passes. In the figure, is appreciated how the fraction of collision slots keeps decreasing once CSMA/ECA$_{\text{Hys+FS}}$ reaches the collision-free schedule.

@@

	\subsubsection{Effect of clock drift over the achieved throughput in saturation}\label{performanceClockDrift}
	Figure~\ref{fig:clockDrift} shows the network aggregated throughput with 16 saturated stations and an increasing clock drift probability.

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/clockDrift/throughput_and_BOS_w_SD-TON.eps}
		\caption{Throughput when increasing the clock drift probability}
		\label{fig:clockDrift}
	\end{figure}

	In Figure~\ref{fig:clockDrift}, a station has a clock drift probability equal to $p_{cd}$. Each station has a probability of $p_{cd}/2$ of miscounting one slot more, and $p_{cd}/2$ of miscounting one slot less. Because CSMA/CA is based on a random backoff, miscounting slots has no significant effect on the throughput. For the CSMA/ECA curve, it is possible to appreciate a slight decrease of the throughput as $p_{cd}$ increases, caused by collisions due to the drift.

	The CSMA/ECA$_{\text{Hys+FS}}$ curve in Figure~\ref{fig:clockDrift} shows instead an increase of the aggregated throughput as $p_{cd}$ grows. Collisions make CSMA/ECA$_{\text{Hys+FS}}$ contenders to increment their backoff stage and aggregate packets for transmissions according to Fair Share, effectively increasing the throughput.

@@

	Emptying the MAC queue in CSMA/ECA$_{\text{Hys+FS}}$ means that nodes will reset their backoff stage and use a random backoff when a new packet arrives at the queue, breaking the collision-free schedule (if any) for CSMA/ECA$_{\text{Hys+FS}}$ contenders. The following show the impact over throughput, delay and time between successful transmissions when using CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ in non-saturated conditions.\\

@@

	In Figure~\ref{fig:unsatResults}a, the aggregated throughput increases linearly for the \emph{CSMA/CA non-saturated} curve until saturation is reached at around 22 nodes, where the throughput begins to degrade. The \emph{CSMA/ECA$_{\text{Hys+FS}}$ non-saturated} curve has a similar behavior, entering saturation at around 60 nodes. Further, at around 30 nodes we see an increase in the average backoff stage for CSMA/ECA$_{\text{Hys+FS}}$ contenders which suggests an increment in collisions. This effect is shown in Figure~\ref{fig:unsatResults}b and Figure~\ref{fig:unsatResults}d, where at around 35 nodes CSMA/ECA$_{\text{Hys+FS}}$ contenders start colliding and dropping packets.

	\textcolor{red}{As indicated by Figure~\ref{fig:unsatResults}b, when $20<N\leq 35$ CSMA/ECA$_{\text{Hys+FS}}$ nodes suffer from an increasing number of  collisions. This is due to nodes entering the contention with a random backoff every time a new packet arrives at their empty MAC queues. Beyond 35 contenders, the MAC queue of CSMA/ECA$_{\text{Hys+FS}}$ nodes starts to fill up, as appreciated in Figure~\ref{fig:unsatResults}f, gradually allowing the construction of longer collision-free schedules due to CSMA/ECA$_{\text{Hys+FS}}$ nodes getting saturated}. This allows CSMA/ECA$_{\text{Hys+FS}}$ to outperform CSMA/CA.\\

@@

	Further, with CSMA/ECA$_{\text{Hys+FS}}$ the percentage of blocked packets from the MAC queue is lower than CSMA/CA or CSMA/CA$_{\text{MaxAg}}$ (see Figure~\ref{fig:blocked-packets}). This is due to the construction of a collision-free schedule which ensures that nodes transmit frequently (Hysteresis) and Fair Share which reduces the average delay given that more packets are transmitted and acknowledged with a single ACK.

@@

	Figure~\ref{fig:unsatResults}c shows the average time between successful transmissions. It is possible to see from the figure that when the saturation point is approached the average time between successful transmissions increases, resembling Figure~\ref{fig:satResults}d. Moreover, this effect is related to an increase in collisions: colliding nodes should begin a new contention for a retransmission, which may result in a successful transmission or another collision, the latter further increasing the average time between successful transmissions.

@@

@@

	Figure~\ref{fig:mixedThroughput-sat} shows the network throughput for different proportions of CSMA/CA nodes in a mixed network setup.

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/throughput-mixed/throughput-saturated-mixed-TON.eps}
		\caption{Network throughput when composed by various proportions of CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes}
		\label{fig:mixedThroughput-sat}
	\end{figure}

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/collisions-mixed/collisions-mixed-saturated-TON.eps}
		\caption{Average percentage of collision slots for the tested mixed network setups proportions}
		\label{fig:mixedCollisions-sat}
	\end{figure}

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/mixed/throughput-mixed/throughput-unsaturated-mixed-TON.eps}
		\caption{Network throughput when composed by various proportions of non-saturated CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes}
		\label{fig:mixedThroughput-unsat}
	\end{figure}

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/mixed/collisions-mixed/collisions-mixed-unsaturated-TON.eps}
		\caption{Average percentage of collision slots for the tested mixed network setups proportions in non-saturated conditions}
		\label{fig:mixCollisions-unsat}
	\end{figure}

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/throughput-per-protocol/throughput-per-protocol-TON.eps}
		\caption{Average station throughput per MAC protocol in a saturated mixed network}
		\label{fig:mixThroughput-sat-perProtocol-TON}
	\end{figure}

	In the figure it is appreciated how the mixed network setups curves lay between the CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ curves. As the proportion of CSMA/CA nodes decreases, the throughput increases as the result of a lower probability of collision, as can be seen in Figure~\ref{fig:mixedCollisions-sat}. A similar behaviour is observed when testing the same proportion of nodes under non-saturated conditions. Figure~\ref{fig:mixedThroughput-unsat} and Figure~\ref{fig:mixCollisions-unsat} show the average aggregated throughput and fraction of collisions slots in a non-saturated mixed network setup.

	As shown in Figure~\ref{fig:mixedThroughput-sat} and Figure~\ref{fig:mixedThroughput-unsat}, at a lower proportion of CSMA/CA nodes ($1/4$) the average aggregated throughput is the greatest among the tested mixed network setups. This is because collisions trigger Hysteresis and Fair Share in CSMA/ECA$_{\text{Hys+FS}}$ nodes, lowering the number of times these nodes enter in a contention and reducing the overall collision probability when compared to an only CSMA/CA network (see Figure~\ref{fig:mixedCollisions-sat} and Figure~\ref{fig:mixCollisions-unsat}). As the proportion of CSMA/CA nodes increases, the network throughput approximates to CSMA/CA.

	\textcolor{red}{Figure~\ref{fig:mixThroughput-sat-perProtocol-TON} shows the average throughput for CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ stations in a saturated mixed setup (half the contenders using CSMA/CA). It is possible to see that for a low total contenders ($N^{*}\leq 12$) CSMA/CA stations attain greater throughput than in a CSMA/CA-only network. This is because in the mixed network setup the other $N^{*}/2$ contenders with CSMA/ECA$_{\text{Hys+FS}}$ use a deterministic backoff, leaving many empty slots between transmissions. This reduces the chances of collisions for CSMA/CA nodes when compared with a CSMA/CA-only network and the same number of total contenders.}

	\textcolor{red}{Still referring to Figure~\ref{fig:mixThroughput-sat-perProtocol-TON}, for $N^{*}>12$ periods of collision-free operation and the aggregation performed with Fair Share allows CSMA/ECA$_{\text{Hys+FS}}$ to have larger channel time than CSMA/CA nodes, which throughput degrades even more than in CSMA/CA-only networks.}

	\textcolor{red}{\subsection{Saturation, channel errors and Schedule Reset}\label{resultsSchedRest}}
	\textcolor{red}{Figure~\ref{fig:throughput-sat-SR} shows the average aggregated throughput of a non-saturated network with a channel error probability, $p_e=0.1$ (see Section~\ref{channelErrorsDef}). We can see that the highest throughput, corresponding to CSMA/ECA$_{\text{Hys+FS}}$ is lower than the one observed in a perfect channel (Figure~\ref{fig:satResults}a). Furthermore, curves with Schedule Reset (SR) seem to have lower aggregated throughput. This is because the backoff stage, and therefore the level of aggregation done by Fair Share, is constantly reduced.}

	\textcolor{red}{Still assessing Figure~\ref{fig:throughput-sat-SR} and focusing on the SR curves, we can see that both \emph{SR aggr. halv. dynStick} curves (with and without Fair Share) have higher aggregated throughput than the rest. This Schedule Reset configuration shows higher aggregated throughput because it reduces the time between successful transmissions (see Figure~\ref{fig:timeBetSxTx-sat-SR}) and mantains collision-free schedules for longer periods of time thanks to the increase in the node's stickiness.}

	\textcolor{red}{Figure~\ref{fig:timeBetSxTx-sat-SR} shows the average time between successful transmissions for the same network setup. Even-though CSMA/ECA$_{\text{Hys+FS}}$ still has a higher metric than CSMA/CA due to the aggregation done by Fair Share, Schedule Reset is able of reducing this metric by almost a $43\%$.}

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/throughput-sat-SR-TON.eps}
		\caption{Average throughput for a saturated network with channel error ($p_e=0.1$). Curves called "aggr" reffer to values of $\gamma=1$, while "halv" refers to a schedule halving (see Section~\ref{aggr} and Algorithm~\ref{alg:schedRest}). \emph{dynStick} reffers to curvers where the node's stickiness is temporarily increased to two after a successful reduction of the schedule}
		\label{fig:throughput-sat-SR}
	\end{figure}

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/timeBetSxTx-sat-SR-TON.eps}
		\caption{Time between successful transmissions for a saturated network with channel error ($p_e=0.1$). Curves called "aggr" reffer to values of $\gamma=1$, while "halv" refers to a schedule halving (see Section~\ref{aggr} and Algorithm~\ref{alg:schedRest}). \emph{dynStick} reffers to curvers where the node's stickiness is temporarily increased to two after a successful reduction of the schedule}
		\label{fig:timeBetSxTx-sat-SR}
	\end{figure}

>>>>

\section{Results}\label{results}
	\begin{figure*}[tb]
		\centering
		\includegraphics[width=0.5\linewidth,angle=-90]{figures/tonFigs/saturation-combined.eps}
		\caption{Saturation traffic results: a) Throughput under saturated conditions; b) Evolution of the fraction of collision slots in a scenario with 70 saturated stations; c) Average percentage of collision slots: fraction of time slots containing collisions; d) Average time between successful transmissions (sx tx), e) Throughput when increasing the clock drift probability}
		\label{fig:satResults}
	\end{figure*}

	%\begin{figure}[tb]
%	\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/clockDrift/throughput_and_BOS_w_SD-TON.eps}
%		\caption{Throughput when increasing the clock drift probability}
%		\label{fig:clockDrift}
%	\end{figure}

	%\begin{figure*}[tb]
%	\centering
%		\subcaptionbox{Throughput under saturated conditions\label{fig:throughput-sat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/throughput-saturated-w-BOS/throughput-saturated-w-BOS-TON.eps}}\hfill
%		\subcaptionbox{Average percentage of collision slots: the fraction of time slots containing collisions\label{fig:collisions-sat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/collisions-saturated/collisions-saturated-TON.eps}}\\
%		\subcaptionbox{Evolution of the fraction of collision slots in a scenario with 70 saturated stations\label{fig:collisions-evolution}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/slots/Pc-evolution-TON.eps}}\hfill
%		\subcaptionbox{Average time between successful transmissions: averaging over all stations\label{fig:serviceTime-sat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/timeBetweenSxTx-sat/timeBetweenSxTx-multiplot-sat-TON.eps}}\\
%		\caption{Saturated traffic results}
%		\label{fig:satTest}
%	\end{figure*}

@@

	This section aims at overviewing the throughput of CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ in saturation, as well as the collision probability, the average time between successful transmissions and the effect of clock drift over the throughput.
	\\
	\subsubsection{Throughput}
	CSMA/ECA$_{\text{Hys+FS}}$ nodes are able to build a collision-free schedule, use the channel more efficiently, and experience a throughput increase as seen in Figure~\ref{fig:satResults}a. Hysteresis allows the allocation of more contenders by increasing the length of the collision-free schedule, while Fair Share ensures an even distribution of the available throughput. This is reflected by the average backoff stage, which value increases with the number of contenders. In contrast, CSMA/CA throughput keeps decreasing due to an augmented number of collisions as the number of nodes increases (see Figure~\ref{fig:satResults}c). Further, Figure~\ref{fig:satResults}b shows the fraction of collision slots for CSMA/ECA$_{\text{Hys+FS}}$ and CSMA/CA as simulation time passes. In the figure, is appreciated how the fraction of collision slots keeps decreasing once CSMA/ECA$_{\text{Hys+FS}}$ reaches collision-free operation.

@@

	\subsubsection{Effect of clock drift over the achieved throughput in saturation}\label{performanceClockDrift}
	Figure~\ref{fig:satResults}e shows the network aggregated throughput with 16 saturated stations and an increasing clock drift probability.

%	\begin{figure}[tb]
%	\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/clockDrift/throughput_and_BOS_w_SD-TON.eps}
%		\caption{Throughput when increasing the clock drift probability}
%		\label{fig:clockDrift}
%	\end{figure}

	In Figure~\ref{fig:satResults}e, a station has a clock drift probability equal to $p_{cd}$. Each station has a probability of $p_{cd}/2$ of miscounting one slot more, and $p_{cd}/2$ of miscounting one slot less. Because CSMA/CA is based on a random backoff, miscounting slots has no significant effect on the throughput. For the CSMA/ECA curve, it is possible to appreciate a slight decrease of the throughput as $p_{cd}$ increases, caused by collisions due to the drift.

	The CSMA/ECA$_{\text{Hys+FS}}$ curve in Figure~\ref{fig:satResults}e shows instead an increase of the aggregated throughput as $p_{cd}$ grows. Collisions make CSMA/ECA$_{\text{Hys+FS}}$ contenders to increment their backoff stage and aggregate packets for transmissions according to Fair Share, effectively increasing the throughput.

@@

	Emptying the MAC queue in CSMA/ECA$_{\text{Hys+FS}}$ means that nodes will reset their backoff stage to zero and use a random backoff when a new packet arrives at the queue, breaking any collision-free operation in CSMA/ECA$_{\text{Hys+FS}}$. The following show the impact over throughput, delay and time between successful transmissions when using CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ in non-saturated conditions.\\

@@

	In Figure~\ref{fig:unsatResults}a, the aggregated throughput increases linearly for the \emph{CSMA/CA} curve until saturation is reached at around 22 nodes, where the throughput begins to degrade. The \emph{CSMA/ECA$_{\text{Hys+FS}}$} curve has a similar behavior, entering saturation at around 60 nodes. Further, at around 30 nodes we see an increase in the average backoff stage for CSMA/ECA$_{\text{Hys+FS}}$ contenders which suggests an increment in collisions. This effect is shown in Figure~\ref{fig:unsatResults}b and Figure~\ref{fig:unsatResults}d, where at around 35 nodes CSMA/ECA$_{\text{Hys+FS}}$ contenders start colliding and dropping packets.

	\textcolor{red}{As indicated by Figure~\ref{fig:unsatResults}b, when $20<N\leq 35$ CSMA/ECA$_{\text{Hys+FS}}$ nodes suffer from an increasing number of  collisions. This is due to nodes emptying their MAC queue quicker due to Fair Share, as shown in Figure~\ref{fig:unsatResults}f and re-entering the contention with a random backoff every time a new packet arrives.}

	\textcolor{red}{This increase in collisions may also cause the dropping of packets due to reaching the maximum retransmission limit. As CSMA/ECA$_{\text{Hys+FS}}$ sends more packets during the same simulation time, it shows a higher fraction of dropped packets in Figure~\ref{fig:unsatResults}d.}

	Beyond 35 contenders, the MAC queue of CSMA/ECA$_{\text{Hys+FS}}$ nodes starts to fill up, gradually allowing longer periods of collision-free operation due to CSMA/ECA$_{\text{Hys+FS}}$ nodes getting saturated. This allows CSMA/ECA$_{\text{Hys+FS}}$ to outperform CSMA/CA.\\

@@

	\textcolor{red}{Further, with CSMA/ECA$_{\text{Hys+FS}}$ the percentage of blocked packets from the MAC queue is lower than CSMA/CA or CSMA/CA$_{\text{MaxAg}}$ (see Figure~\ref{fig:blocked-packets}). This is due to the construction of collision-free schedules which ensure that nodes transmit frequently (Hysteresis) and Fair Share which reduces the average delay given that more packets are transmitted and acknowledged with a single ACK (A-MPDU aggregation).}

@@

	Figure~\ref{fig:unsatResults}c shows the average time between successful transmissions. It is possible to see from the figure that when CSMA/ECA$_{\text{Hys+FS}}$ approaches the saturation point the average time between successful transmissions increases, resembling Figure~\ref{fig:satResults}d.

	%Moreover, this effect is related to an increase in collisions: colliding nodes should begin a new contention for a retransmission, which may result in a successful transmission or another collision, the latter further increasing the average time between successful transmissions.

@@

	\begin{figure*}[tb]
		\centering
		\includegraphics[width=0.5\linewidth,angle=-90]{figures/tonFigs/coexistence-combined.eps}
		\caption{Coexistence results: a) Network throughput when composed by various proportions of CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes; b) Average percentage of collision slots for the tested mixed network setups proportions; c) Network throughput when composed by various proportions of non-saturated CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes; d) Average percentage of collision slots for the tested mixed network setups proportions in non-saturated conditions; e) Average station throughput per MAC protocol in a saturated mixed network}
		\label{fig:coexResults}
	\end{figure*}

@@

	Figure~\ref{fig:coexResults}a shows the network throughput for different proportions of CSMA/CA nodes in a mixed network setup.

%	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/throughput-mixed/throughput-saturated-mixed-TON.eps}
%		\caption{Network throughput when composed by various proportions of CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes}
%		\label{fig:mixedThroughput-sat}
%	\end{figure}
%	
%	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/collisions-mixed/collisions-mixed-saturated-TON.eps}
%		\caption{Average percentage of collision slots for the tested mixed network setups proportions}
%		\label{fig:mixedCollisions-sat}
%	\end{figure}
%	
%	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/mixed/throughput-mixed/throughput-unsaturated-mixed-TON.eps}
%		\caption{Network throughput when composed by various proportions of non-saturated CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ nodes}
%		\label{fig:mixedThroughput-unsat}
%	\end{figure}
%	
%	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/mixed/collisions-mixed/collisions-mixed-unsaturated-TON.eps}
%		\caption{Average percentage of collision slots for the tested mixed network setups proportions in non-saturated conditions}
%		\label{fig:mixCollisions-unsat}
%	\end{figure}
%	
%	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/mixed/throughput-per-protocol/throughput-per-protocol-TON.eps}
%		\caption{Average station throughput per MAC protocol in a saturated mixed network}
%		\label{fig:mixThroughput-sat-perProtocol-TON}
%	\end{figure}

	In the figure it is appreciated how the mixed network setups curves lay between the CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ curves. As the proportion of CSMA/CA nodes decreases, the throughput increases as the result of a lower probability of collision, as can be seen in Figure~\ref{fig:coexResults}b. A similar behaviour is observed when testing the same proportion of nodes under non-saturated conditions. Figure~\ref{fig:coexResults}c and Figure~\ref{fig:coexResults}d show the average aggregated throughput and fraction of collisions slots in a non-saturated mixed network setup.

	As shown in Figure~\ref{fig:coexResults}a and Figure~\ref{fig:coexResults}c, at a lower proportion of CSMA/CA nodes ($1/4$) the average aggregated throughput is the greatest among the tested mixed network setups. This is because collisions trigger Hysteresis and Fair Share in CSMA/ECA$_{\text{Hys+FS}}$ nodes, lowering the number of times these nodes enter in a contention and reducing the overall collision probability when compared to an only CSMA/CA network (see Figure~\ref{fig:coexResults}b, Figure~\ref{fig:coexResults}d and Figure~\ref{fig:coexResults}e). As the proportion of CSMA/CA nodes increases, the network throughput approximates to CSMA/CA.

	\textcolor{red}{Figure~\ref{fig:coexResults}e shows the average throughput for CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ stations in a saturated mixed setup (half the contenders using CSMA/CA). It is possible to see that for a low total contenders ($N^{*}\leq 12$) CSMA/CA stations attain greater throughput than in a CSMA/CA-only network. Again, this is because in the mixed network setup the other $N^{*}/2$ contenders with CSMA/ECA$_{\text{Hys+FS}}$ use a deterministic backoff, leaving many empty slots between transmissions.}

	\textcolor{red}{Still referring to Figure~\ref{fig:coexResults}e, for $N^{*}>12$ periods of collision-free operation and the aggregation performed with Fair Share allows CSMA/ECA$_{\text{Hys+FS}}$ to have larger channel time than CSMA/CA nodes, which throughput degrades even more than in CSMA/CA-only.}

	%####################################################################################

	\textcolor{red}{\subsection{Channel errors and Schedule Reset}\label{resultsSchedRest}}

	\begin{figure*}[tb]
		\centering
		\includegraphics[width=0.65\linewidth,angle=-90]{figures/tonFigs/schedReset-combined.eps}
		\caption{Performance results with Schedule Reset. Curves called "aggr" reffer to values of $\gamma=1$, while "halv" refers to a schedule halving (see Section~\ref{aggr} and Algorithm~\ref{alg:schedRest}). \emph{dynStick} reffers to curvers where the node's stickiness is temporarily increased to two after a successful reduction of the schedule: a) Average throughput for a saturated network; b) Average throughput for a non-saturated network; c) Time between successful transmissions for a saturated network; d) Time between successful transmissions for a non-saturated network}
		\label{fig:SchedResetResults}
	\end{figure*}

	\textcolor{red}{Figure~\ref{fig:SchedResetResults}a shows the average aggregated throughput of a saturated network with a channel error probability, $p_e=0.1$ (see Section~\ref{channelErrorsDef}). We can see that the highest throughput, corresponding to CSMA/ECA$_{\text{Hys+FS}}$ is lower than the one observed in a perfect channel (Figure~\ref{fig:satResults}a). Furthermore, curves with Schedule Reset (SR) seem to have lower aggregated throughput. This is because the backoff stage, and therefore the level of aggregation done by Fair Share, is repeatedly reduced.}

	\textcolor{red}{Still assessing Figure~\ref{fig:SchedResetResults}a and focusing on the SR curves, we can see that both \emph{SR aggr. halv. dynStick} curves (with and without Fair Share) have higher aggregated throughput than the rest. This Schedule Reset configuration reduces the time between successful transmissions (see Figure~\ref{fig:SchedResetResults}c) and mantains collision-free operation for longer periods of time thanks to the increase in the node's stickiness.}

	\textcolor{red}{Figure~\ref{fig:SchedResetResults}c shows the average time between successful transmissions for the same network setup. Even-though CSMA/ECA$_{\text{Hys+FS}}$ still has a higher metric than CSMA/CA due to the aggregation done by Fair Share, Schedule Reset is able of reducing this metric by almost a $43\%$.}

	\textcolor{red}{Figure~\ref{fig:SchedResetResults}b and Figure~\ref{fig:SchedResetResults}d show the aggregate througput and time between successful transmissions, respectively for a non-saturated network with the same $p_e=0.1$. Although a reduction in time between successfull transmissions is achieved with Schedule Reset, under non-saturated conditions CSMA/ECA$_{\text{Hys+FS}}$ nodes reset their schedule every time they empty their MAC queues. Therefore Schedule Reset's benefits are relevant only when in saturation.}

	%\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/throughput-sat-SR-TON.eps}
%		\caption{Average throughput for a saturated network with channel error ($p_e=0.1$). Curves called "aggr" reffer to values of $\gamma=1$, while "halv" refers to a schedule halving (see Section~\ref{aggr} and Algorithm~\ref{alg:schedRest}). \emph{dynStick} reffers to curvers where the node's stickiness is temporarily increased to two after a successful reduction of the schedule}
%		\label{fig:throughput-sat-SR}
%	\end{figure}
%	
%	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/timeBetSxTx-sat-SR-TON.eps}
%		\caption{Time between successful transmissions for a saturated network with channel error ($p_e=0.1$). Curves called "aggr" reffer to values of $\gamma=1$, while "halv" refers to a schedule halving (see Section~\ref{aggr} and Algorithm~\ref{alg:schedRest}). \emph{dynStick} reffers to curvers where the node's stickiness is temporarily increased to two after a successful reduction of the schedule}
%		\label{fig:timeBetSxTx-sat-SR}
%	\end{figure}
